Exploring dimensions of biodiversity in Japanese ferns

### 2020-07-13 presentations/bsa_2020-07-27/index.Rmd
Update slides

```{r load-packages, cache = FALSE}
library(magick)
# library(gt)
library(ggtext)
library(drake)
library(knitr)
library(patchwork)
library(scales)
library(scico)
library(magrittr)
library(broom)
library(assertr)
library(sf)
library(rcartocolor)
library(tidyverse)
```

@@

@@

# Write out here, then crop manually
ggsave(
  plot = title_image, 
  filename = here::here("presentations/bsa_2020-07-27/images/title.png"),
  dpi = 300)

```

@@

@@

status_labels <-
protected_areas %>%
  as_tibble %>%
  select(status, area) %>%
  group_by(status) %>%
  summarize(
    area = sum(area),
    .groups = "drop"
  ) %>%
  mutate(
    area = as.numeric(area),
    area_chr = area %>% round %>% scales::number(big.mark = ","),
    area_percent = area %>% magrittr::divide_by(377900) %>% scales::percent(),
    label = glue::glue("{str_to_sentence(status)} ({area_percent})") %>% set_names(status)
) %>%
  pull(label)

ggplot() +
  geom_sf(data = japan) +
  geom_sf(data = protected_areas, aes(fill = status), color = "transparent") +
  coord_sf(xlim = c(japan_box$xmin, japan_box$xmax), ylim = c(japan_box$ymin, japan_box$ymax)) +
  labs(
    fill = "Conservation\nstatus"
  ) +
  scale_fill_viridis_d(breaks = names(status_labels), labels = status_labels)
```

@@

# plot showing protected status of high richness areas
richness_initial_plot <- ggplot() +
  geom_sf(data = japan_cropped_to_richness) +
  geom_sf(data = protected_richness, aes(fill = status)) + 
  scale_fill_viridis_d(breaks = c("low", "medium", "high"), labels = c("Low", "Medium", "High")) +
  labs(
    title = "Top 5% richness",
    fill = "Conservation\nstatus"
  )

@@

# plot showing protected status of high PE areas
pe_initial_plot <- ggplot() +
  geom_sf(data = japan_cropped_to_pe) +
  geom_sf(data = protected_pe, aes(fill = status)) + 
  scale_fill_viridis_d(breaks = c("low", "medium", "high"), labels = c("Low", "Medium", "High"), na.value = "white") +
  labs(
    title = "Significantly high PE",
    fill = "Conservation\nstatus"
  )

>>>>

```{r load-packages, cache = FALSE}
library(conflicted)
library(magick)
library(ggtext)
library(drake)
library(knitr)
library(patchwork)
library(scales)
library(scico)
library(magrittr)
library(broom)
library(assertr)
library(sf)
library(rcartocolor)
library(tidyverse)

# Resolve conflicts
conflicted::conflict_prefer("map", "purrr")
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("gather", "tidyr")
conflicted::conflict_prefer("extract", "magrittr")

# Specify custom themes
simple_map_theme <- function () {
  theme(
    panel.grid.major = element_line(color = "white", size = 0.1),
    panel.background = element_rect(fill = "grey70"),
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
}

trend_line_theme <- function () {
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  )
}
```

@@

# Change phyloregion cluster to factor
ses_div_ferns_spatial <- ses_div_ferns_spatial %>%
  mutate(taxonomic_cluster = as.factor(taxonomic_cluster) %>% fct_infreq %>% as.numeric %>% as.factor) %>%
  mutate(phylo_cluster = as.factor(phylo_cluster) %>% fct_infreq %>% as.numeric %>% as.factor)

@@

# Write out here, then crop manually
ggsave(
  plot = title_image, 
  filename = here::here("presentations/bsa_2020-07-27/images/title.png"),
  dpi = 300)
```

@@

---
## Species richness is humped-shaped

```{r lat-bins}
# Bin SES diversity results by latitudinal band (0.2 degree width)
ses_div_binned <-
  ses_div_ferns_spatial %>%
  mutate(
    longitude = sf::st_centroid(geometry) %>% st_coordinates %>% magrittr::extract(,1),
    latitude = sf::st_centroid(geometry) %>% st_coordinates %>% magrittr::extract(,2)) %>%
  as_tibble %>%
  dplyr::select(-geometry) %>%
  mutate(
    latitude = cut_width(latitude, width = 0.2, closed = "left") %>%
      str_split(",") %>%
      map_chr(1) %>%
      parse_number) %>%
  group_by(latitude) %>%
  summarize(
    across(c(richness, pe_obs_rank, contains("obs_z")), ~mean(.x, na.rm = TRUE))
  )

# Get min and max latitude for setting common limits between
# maps and trendlines
ja_lat_min <- ses_div_binned %>% pull(latitude) %>% min()
ja_lat_max <- ses_div_binned %>% pull(latitude) %>% max()
```

```{r richness, out.width = "120%"}
# Fern richness
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = richness)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(fill = "No. taxa") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  ) +
  ylim(ja_lat_min, ja_lat_max)

# - line tracking mean richness
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = richness)) +
  geom_path() +
  labs(
    x = "No. taxa"
  ) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  ) +
  ylim(ja_lat_min, ja_lat_max)

a + b + plot_layout(widths = c(3,1), heights = c(1,1))
```

---
## Other raw values are correlated with richness

```{r raw-div, out.width = "120%", fig.asp = 0.7}
# Richness
a <- ggplot(ses_div_ferns_spatial, aes(fill = richness)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Taxonomic richness") +
  simple_map_theme()

b <- ggplot(ses_div_ferns_spatial, aes(fill = pd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Phylogenetic diversity") +
  simple_map_theme()

c <- ggplot(ses_div_ferns_spatial, aes(fill = fd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Functional diversity") +
  simple_map_theme()

d <- ggplot(ses_div_ferns_spatial, aes(x = richness, y = pd_obs)) +
  geom_point(shape = 1) +
  # Fit a logarithmic general additive model
  # https://stackoverflow.com/questions/40711980/r-ggplot2-fit-curve-to-scatter-plot
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(log(x)), size = 0.5) +
  labs(
    x = "No. taxa",
    y = "PD"
  ) +
  jntools::standard_theme()

e <- ggplot(ses_div_ferns_spatial, aes(x = richness, y = fd_obs)) +
  geom_point(shape = 1) +
  geom_smooth(se = FALSE, method = "gam", formula = y ~ s(log(x)), size = 0.5) +
  labs(
    x = "No. taxa",
    y = "FD"
  ) +
  jntools::standard_theme()

a + b + c +
  grid::textGrob("...as expected") + d + e +
  plot_layout(nrow = 2, heights = c(0.5, 0.5))
```

---
## Compare raw values to random expectation with <u>**S**</u>tandard <u>**E**</u>ffect <u>**S**</u>ize

.pull-left[
$SES = \frac{x - mean(x)}{sd(x)}$
]

.pull-right[
```{r ses-example}
null <- tibble(value = rnorm(10000))

obs_1 <- 0.95
obs_2 <- 0.5
obs_3 <- 0.05

ggplot(null, aes(x = value)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0)
```
]

---
## Compare raw values to random expectation with <u>**S**</u>tandard <u>**E**</u>ffect <u>**S**</u>ize

PD is mostly clustered throughout

```{r ses-pd, out.width = "120%"}
# SES PD
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = pd_obs_z)) +
  geom_sf(color = "transparent") +
  labs(fill = "SES PD") +
  scale_fill_scico(
    palette = "romaO",
    na.value="grey80",
    direction = -1,
    limits = c(
      -get_limit(ses_div_ferns_spatial, pd_obs_z, "abs"),
      get_limit(ses_div_ferns_spatial, pd_obs_z, "abs")
    )) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  )

# - line tracking mean SES RPD
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = pd_obs_z)) +
  geom_path() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(
    x = "SES PD"
  ) +
  ylim(24, 46) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  )

a + b + plot_layout(widths = c(3,1), heights = c(1,1))
```

---
## Compare branch lengths with <u>**R**</u>elative <u>**P**</u>hylogenetic <u>**D**</u>iversity

RPD = original branch lengths / all branch lengths equal

---
## Compare branch lengths with <u>**R**</u>elative <u>**P**</u>hylogenetic <u>**D**</u>iversity

RPD is high in the N and S

```{r ses-rpd, out.width = "120%"}
# SES RPD
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = rpd_obs_z)) +
  geom_sf(color = "transparent") +
  labs(fill = "SES RPD") +
  scale_fill_scico(
    palette = "romaO",
    na.value="grey80",
    direction = -1,
    limits = c(
      -get_limit(ses_div_ferns_spatial, rpd_obs_z, "abs"),
      get_limit(ses_div_ferns_spatial, rpd_obs_z, "abs")
    )) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  )

# - line tracking mean SES RPD
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = rpd_obs_z)) +
  geom_path() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(
    x = "SES RPD"
  ) +
  ylim(24, 46) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  )

a + b + plot_layout(widths = c(3,1), heights = c(1,1))
```

---
## <u>**R**</u>elative <u>**F**</u>unctional <u>**D**</u>iversity: functional analog of RPD

RFD also high in the south

```{r ses-rpd-rfd, out.width = "120%"}
# SES RPD
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = rfd_obs_z)) +
  geom_sf(color = "transparent") +
  labs(fill = "SES RFD") +
  scale_fill_scico(
    palette = "romaO",
    na.value="grey80",
    direction = -1,
    limits = c(
      -get_limit(ses_div_ferns_spatial, rfd_obs_z, "abs"),
      get_limit(ses_div_ferns_spatial, rfd_obs_z, "abs")
    )) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  )

# - line tracking mean SES RPD
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = rfd_obs_z)) +
  geom_path() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(
    x = "SES RFD"
  ) +
  ylim(24, 46) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    plot.margin = margin(l = -1, unit = "inch")
  )

a + b + plot_layout(widths = c(3,1), heights = c(1,1))
```

---
## Phylogenetic endemism: PD weighted by range size

Southern islands are hotspots of endemism

```{r pe, out.width = "120%"}
# PE
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = pe_obs_rank)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(fill = "PE Rank") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  ylim(ja_lat_min, ja_lat_max) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  )

# - line tracking mean richness
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = pe_obs_rank)) +
  geom_path() +
  labs(
    x = "PE Rank"
  ) +
  ylim(ja_lat_min, ja_lat_max) +
  trend_line_theme()

a + b + plot_layout(widths = c(3,1), heights = c(1,1))
```

---
## Bioregions

Most sites fall into four bioregions

```{r tax-bioregion, out.width = "120%", fig.asp = 0.7}
ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_light()
```

@@

status_labels <-
protected_areas %>%
  as_tibble %>%
  dplyr::select(status, area) %>%
  group_by(status) %>%
  summarize(
    area = sum(area),
    .groups = "drop"
  ) %>%
  mutate(
    area = as.numeric(area),
    area_chr = area %>% round %>% scales::number(big.mark = ","),
    area_percent = area %>% magrittr::divide_by(377900) %>% scales::percent(),
    label = glue::glue("{str_to_sentence(status)} ({area_percent})") %>% set_names(status)
) %>%
  pull(label)

ggplot() +
  geom_sf(data = japan) +
  geom_sf(data = protected_areas, aes(fill = status), color = "transparent") +
  coord_sf(xlim = c(japan_box$xmin, japan_box$xmax), ylim = c(japan_box$ymin, japan_box$ymax)) +
  labs(
    fill = "Conservation\nstatus"
  ) +
  scale_fill_viridis_d(breaks = names(status_labels), labels = status_labels) +
  theme(
    axis.ticks = element_blank()
  )
```

@@

# plot showing protected status of high richness areas
richness_initial_plot <- ggplot() +
  geom_sf(data = japan_cropped_to_richness) +
  geom_sf(data = protected_richness, aes(fill = status)) + 
  scale_fill_viridis_d(breaks = c("low", "medium", "high"), labels = c("Low", "Medium", "High")) +
  labs(
    title = "Top 5% richness",
    fill = "Conservation\nstatus"
  ) +
  theme(
    axis.ticks = element_blank()
  )

@@

# plot showing protected status of high PE areas
pe_initial_plot <- ggplot() +
  geom_sf(data = japan_cropped_to_pe) +
  geom_sf(data = protected_pe, aes(fill = status)) + 
  scale_fill_viridis_d(breaks = c("low", "medium", "high"), labels = c("Low", "Medium", "High"), na.value = "white") +
  labs(
    title = "Significantly high PE",
    fill = "Conservation\nstatus"
  ) +
  theme(
    axis.ticks = element_blank()
  )
